---
title: "POMA: Statistical analysis tool for omics data"
subtitle: "Exploratory Data Analysis Report"
date: '`r format(Sys.Date(), "%B, %Y")`'
output:
  prettydoc::html_pretty:
    toc: true
    theme: cayman
    highlight: github
params:
  n: NA
  t: NA
---

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
# This file is part of POMA.

# POMA is free software: you can redistribute it and/or modify
# it under the terms of the GNU General Public License as published by
# the Free Software Foundation, either version 3 of the License, or
# (at your option) any later version.

# POMA is distributed in the hope that it will be useful,
# but WITHOUT ANY WARRANTY; without even the implied warranty of
# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
# GNU General Public License for more details.

# You should have received a copy of the GNU General Public License
# along with POMA. If not, see <https://www.gnu.org/licenses/>.

data <- params$n
target <- params$t

# data <- read.table("/Users/pol/Dropbox/JosepVillanueva/231eribulin-counts.txt", header = T, sep = "\t")
# target <- read.table("/Users/pol/Dropbox/JosepVillanueva/231eribulin-samples.txt", header = T, sep = "\t")
data <- read.table("/home/pol/Escritorio/R problems/231eribulin-counts.txt", header = T, sep = "\t")
target <- read.table("/home/pol/Escritorio/R problems/231eribulin-samples.txt", header = T, sep = "\t")
```
 
```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
colnames(target) <- c("Sample", "Group", "Batch")

prot_names <- data$Accession
remove <- ncol(data) - nrow(target)
data <- data[, -c(1:remove)]
# counts <- apply(data, 2, function(x) as.numeric(as.character(x)))
data <- data.frame(t(data))
colnames(data) <- prot_names
```

# Summary

  + Your data has **`r nrow(target)`** subjects, **`r ncol(data)`** proteines and **`r length(table(target$Group))`** groups, that are **`r noquote(paste(shQuote(levels(as.factor(target$Group))), collapse=", "))`**.    

  + A **`r round(sum(is.na(data))/(nrow(data)*(ncol(data)))*100, 2)`%** of the values in your data are NAs (missing values). 

## Subject count by Groups

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
library(ggplot2)
library(reshape)

counts <- data.frame(table(target$Group))
colnames(counts) <- c("Group", "Counts")

ggplot(counts, aes(Group, Counts, fill = Group)) +
  geom_bar(stat="identity") +
  theme_minimal() +
  scale_fill_brewer(palette = "Dark2")
```
  
# Barplot scaled to the median

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
proteines <- params$n
target <- params$t

### 

colnames(target) <- c("Sample", "Treatment", "Batch")

rownames(proteines) <- proteines$Accession
remove <- ncol(proteines) - nrow(target)
proteines <- proteines[, -c(1:remove)]

prot_names <- rownames(proteines)
counts <- apply(proteines, 2, function(x) as.numeric(as.character(x)))

## PRE-PROCESSING (pp.msms.data)

if (sum(is.na(counts))) {
  counts[is.na(counts)] <- 0
}

fl1 <- apply(counts, 1, function(x) sum(x) > 0)

fl2 <- substring(prot_names, nchar(prot_names) - 1) == "-R"
flags <- (!fl2 & fl1)
counts <- counts[flags, ]

###

tcnts <- apply(counts, 2, sum)
medt <- median(tcnts)
div <- data.frame(median = tcnts/medt)
div$sample <- rownames(div)
div$Group <- target$Treatment

ggplot(div, aes(x = sample, y = median, fill = Group)) +
  geom_bar(stat="identity")+
  theme_minimal() +
  xlab("") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_fill_brewer(palette = "Dark2") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")

```

# Normalization plot

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
spcm2 <- batch.neutralize(counts, target$Batch, half = TRUE, sqrt.trans = TRUE)

total <- as.data.frame(t(spcm2))
colnames(total) <- prot_names

total <- cbind(target[, 1:2], total)
colnames(total)[1:2] <- c("Sample", "Group")

### NORMALIZATION

normtable_subjects <- total %>% reshape2::melt()

normtable_subjects %>% 
  ggplot(aes(Sample, log2(value + 1), color = Group)) + 
  geom_jitter() + 
  geom_boxplot(color = "black", outlier.colour = NA) + 
  theme(legend.position = "none") + 
  theme_minimal() + 
  xlab("") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "top") +
  scale_color_brewer(palette = "Dark2")
```

# Density plot

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
normtable_subjects %>% 
  ggplot(aes(x = value)) +
  geom_density(aes(group = Sample, color = Group), alpha = 0.4) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_color_brewer(palette = "Dark2")
```

# Principal Component Analysis

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
res_pca1 <- mixOmics::pca(total[, 3:ncol(total)])
res_pca <- cbind(target, res_pca1$x)

ggplot(res_pca, aes(PC1, PC2, color = Treatment, shape = Batch)) +
  geom_point(size = 3, alpha = 0.8) +
  ggrepel::geom_label_repel(aes(label = Sample), show.legend = F) +
  theme_minimal() +
  xlab(paste0("PC1 (", round(100*(res_pca1$explained_variance)[1], 2), "%)")) +
  ylab(paste0("PC2 (", round(100*(res_pca1$explained_variance)[2], 2), "%)")) +
  scale_color_brewer(palette = "Dark2")
```

# Cluster by Group

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
dend1 <- total[, 3:ncol(total)] %>% 
  dist %>% 
  hclust 
order <- dend1$order
target2 <- target[order, ]
my_group <- as.numeric(as.factor(target2$Treatment))
colSide <- brewer.pal(8, "Dark2")[my_group]

dend1 <- dend1 %>%
  as.dendrogram() %>%
  set("labels_colors", colSide) %>%
plot()
```

# Cluster by Batch

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
dend2 <- total[, 3:ncol(total)] %>% 
  dist %>% 
  hclust 
order <- dend2$order
target2 <- target[order, ]
my_group <- as.numeric(as.factor(target2$Batch))
colSide <- brewer.pal(8, "Dark2")[my_group]

dend2 <- dend2 %>%
  as.dendrogram() %>%
  set("labels_colors", colSide) %>%
  plot()
```

# Heatmap

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
my_group <- as.numeric(as.factor(target$Treatment))
colSide <- brewer.pal(8, "Dark2")[my_group]
colMain <- colorRampPalette( c("green", "black", "red"), space = "rgb")(64)

heatmap(t(total[, 3:ncol(total)]), ColSideColors = colSide, col = colMain, labRow = NA)
```

# Important Features

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
spc_scatterplot <- function(counts, treat, trans = "log2", minSpC = 2, minLFC = 1){
    
    treat <- as.factor(treat)
    mspc <- data.frame(t(apply(counts, 1, function(x) tapply(x, treat, mean))))
    mx <- apply(mspc[, 1:2], 1, max)
    mn <- apply(mspc[, 1:2], 1, min)
    fl <- apply(mspc[, 1:2], 1, function(x) max(x) >= minSpC) & abs(log2(mx/mn)) >= minLFC
    
    if (trans == "none") {
      
      s_mspc <- mspc[fl ,]
      
      p <- ggplot(mspc, aes(mspc[,1], mspc[,2])) +
        geom_point() +
        geom_point(s_mspc, mapping = aes(x = s_mspc[,1], y = s_mspc[,2], color = "red")) +
        geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
        geom_abline(intercept = 0, slope = 2, linetype = "dashed", color = "navy") +
        geom_abline(intercept = 0, slope = 0.5, linetype = "dashed", color = "navy") +
        xlab(paste("mean SpC(", colnames(mspc)[1], ")", sep = "")) +
        ylab(paste("mean SpC(", colnames(mspc)[2], ")", sep = "")) +
        theme_minimal() +
        theme(legend.position = "none")
    }
    
    if (trans == "sqrt") {
      
      sqrtm <- sqrt(mspc)
      
      s_mspc <- sqrtm[fl ,]
      
      p <- ggplot(sqrtm, aes(sqrtm[,1], sqrtm[,2])) +
        geom_point() +
        geom_point(s_mspc, mapping = aes(x = s_mspc[,1], y = s_mspc[,2], color = "red")) +
        geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
        geom_abline(intercept = 0, slope = sqrt(2), linetype = "dashed", color = "navy") +
        geom_abline(intercept = 0, slope = 1/sqrt(2), linetype = "dashed", color = "navy") +
        xlab(paste("root mean SpC(", colnames(mspc)[1], ")", sep = "")) +
        ylab(paste("root mean SpC(", colnames(mspc)[2], ")", sep = "")) +
        theme_minimal() +
        theme(legend.position = "none")
    }
    
    if (trans == "log2") {
      
      logm <- log2(mspc + 0.01)
      mxx <- max(logm[, 1])
      mxy <- max(logm[, 2])
      mx <- max(mxx, mxy)
      
      s_mspc <- logm[fl ,]
      
      p <- ggplot(logm, aes(logm[,1], logm[,2])) +
        geom_point() +
        geom_point(s_mspc, mapping = aes(x = s_mspc[,1], y = s_mspc[,2], color = "red")) +
        geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
        geom_abline(intercept = log2(2), slope = 1, linetype = "dashed", color = "navy") +
        geom_abline(intercept = -log2(2), slope = 1, linetype = "dashed", color = "navy") +
        xlab(paste("log2(mean SpC(", colnames(mspc)[1], "))", sep = "")) +
        ylab(paste("log2(mean SpC(", colnames(mspc)[2], "))", sep = "")) +
        theme_minimal() +
        theme(legend.position = "none")
    }
    
    return(p)
  }
  
spc_scatterplot(spcm2, target$Treatment, trans = "sqrt", minSpC = 2, minLFC = 2)
```

