---
title: "POMA: Statistical analysis tool for omics data"
subtitle: "Exploratory Data Analysis Report"
date: '`r format(Sys.Date(), "%B, %Y")`'
output: 
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
params:
  raw: NA
  norm: NA
  corrected: NA
---

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
# This file is part of POMA.

# POMA is free software: you can redistribute it and/or modify
# it under the terms of the GNU General Public License as published by
# the Free Software Foundation, either version 3 of the License, or
# (at your option) any later version.

# POMA is distributed in the hope that it will be useful,
# but WITHOUT ANY WARRANTY; without even the implied warranty of
# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
# GNU General Public License for more details.

# You should have received a copy of the GNU General Public License
# along with POMA. If not, see <https://www.gnu.org/licenses/>.

raw <- params$raw
norm <- params$norm
corrected <- params$corrected
```

# Summary

## Raw Data

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
raw
```

## Normalized Data

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
norm
```

## Normalized and Batch Corrected Data

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
corrected
```

# Median Bar Plot

## Raw

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
### BARPLOT - SCALE TO THE MEDIAN RAW
  
counts <- Biobase::exprs(raw)

tcnts <- apply(counts, 2, sum)
medt <- median(tcnts)
div <- data.frame(median = tcnts/medt) 
div$sample <- rownames(div)
div$Treatment <- target$Treatment

ggplot(div, aes(x = sample, y = median, fill = Treatment)) +
  geom_bar(stat="identity")+
  theme_minimal() +
  xlab("") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_fill_brewer(palette = "Dark2") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")
```

## Normalized

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
### BARPLOT - SCALE TO THE MEDIAN NORMALIZED

counts <- Biobase::exprs(norm)

tcnts <- apply(counts, 2, sum)
medt <- median(tcnts)
div <- data.frame(median = tcnts/medt) 
div$sample <- rownames(div)
div$Treatment <- target$Treatment

##

ggplot(div, aes(x = sample, y = median, fill = Treatment)) +
  geom_bar(stat="identity")+
  theme_minimal() +
  xlab("") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_fill_brewer(palette = "Dark2") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")
```

## Normalized and Batch Corrected

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
### BARPLOT - SCALE TO THE MEDIAN CORRECTED

counts <- Biobase::exprs(corrected)

tcnts <- apply(counts, 2, sum)
medt <- median(tcnts)
div <- data.frame(median = tcnts/medt) 
div$sample <- rownames(div)
div$Treatment <- target$Treatment

##

ggplot(div, aes(x = sample, y = median, fill = Treatment)) +
  geom_bar(stat="identity")+
  theme_minimal() +
  xlab("") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_fill_brewer(palette = "Dark2") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")
```

# Normalization Boxplots

## Raw

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
#### RAW

target <- pData(raw)
total <- as.data.frame(t(exprs(raw)))
total <- cbind(Sample = rownames(target), Treatment = target$Treatment, total)

normtable_subjectsX <- total %>% reshape2::melt()

normtable_subjects1 <- normtable_subjectsX %>%
  filter(value >= 2)

normtable_subjects1 %>%
  ggplot(aes(Sample, log2(value), color = Treatment)) +
  geom_jitter() +
  geom_boxplot(color = "black", outlier.colour = NA) +
  theme(legend.position = "none") +
  theme_minimal() +
  xlab("") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top") +
  scale_color_brewer(palette = "Dark2")
```

## Normalized

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
#### NORMALIZED

target <- pData(norm)
total2 <- as.data.frame(t(exprs(norm)))
total2 <- cbind(Sample = rownames(target), Treatment = target$Treatment, total2)

normtable_subjectsXX <- total2 %>% reshape2::melt()

normtable_subjects2 <- normtable_subjectsXX %>%
  filter(value >= 2)

normtable_subjects2 %>% 
  ggplot(aes(Sample, log2(value), color = Treatment)) + 
  geom_jitter() + 
  geom_boxplot(color = "black", outlier.colour = NA) + 
  theme(legend.position = "top") + 
  theme_minimal() + 
  xlab("") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "top") +
  scale_color_brewer(palette = "Dark2")
```

## Normalized and Batch Corrected

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
#### CORRECTED

target <- pData(corrected)
total3 <- as.data.frame(t(exprs(corrected)))
total3 <- cbind(Sample = rownames(target), Treatment = target$Treatment, total3)

normtable_subjectsXXX <- total3 %>% reshape2::melt()

normtable_subjects3 <- normtable_subjectsXXX %>%
  filter(value >= 2)

normtable_subjects3 %>% 
  ggplot(aes(Sample, log2(value), color = Treatment)) + 
  geom_jitter() + 
  geom_boxplot(color = "black", outlier.colour = NA) + 
  theme(legend.position = "top") + 
  theme_minimal() + 
  xlab("") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "top") +
  scale_color_brewer(palette = "Dark2")
```

# Density Plots

## Raw

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
#### RAW

normtable_subjectsX %>% 
  ggplot(aes(x = value)) +
  geom_density(aes(group = Sample, color = Treatment), alpha = 0.4) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_color_brewer(palette = "Dark2") +
  xlim(0, 30)
```

## Normalized

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
#### NORMALIZED

normtable_subjectsXX %>% 
  ggplot(aes(x = value)) +
  geom_density(aes(group = Sample, color = Treatment), alpha = 0.4) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_color_brewer(palette = "Dark2") +
  xlim(0, 30)
```

## Normalized and Batch Corrected

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
#### CORRECTED

normtable_subjectsXXX %>% 
  ggplot(aes(x = value)) +
  geom_density(aes(group = Sample, color = Treatment), alpha = 0.4) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_color_brewer(palette = "Dark2") +
  xlim(0, 30)
```

# Principal Component Analysis

## Raw

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
#### RAW

res_pca1 <- mixOmics::pca(t(exprs(raw)))
res_pca <- cbind(pData(raw), res_pca1$x)
  
ggplot(res_pca, aes(PC1, PC2, color = Treatment, shape = Batch)) +
  geom_point(size = 3, alpha = 0.8) +
  ggrepel::geom_label_repel(aes(label = rownames(res_pca)), show.legend = F) +
  theme_minimal() +
  xlab(paste0("PC1 (", round(100*(res_pca1$explained_variance)[1], 2), "%)")) +
  ylab(paste0("PC2 (", round(100*(res_pca1$explained_variance)[2], 2), "%)")) +
  scale_color_brewer(palette = "Dark2")
```

## Normalized

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
#### NORMALIZED

res_pca1 <- mixOmics::pca(t(exprs(norm)))
res_pca <- cbind(pData(norm), res_pca1$x)
  
ggplot(res_pca, aes(PC1, PC2, color = Treatment, shape = Batch)) +
  geom_point(size = 3, alpha = 0.8) +
  ggrepel::geom_label_repel(aes(label = rownames(res_pca)), show.legend = F) +
  theme_minimal() +
  xlab(paste0("PC1 (", round(100*(res_pca1$explained_variance)[1], 2), "%)")) +
  ylab(paste0("PC2 (", round(100*(res_pca1$explained_variance)[2], 2), "%)")) +
  scale_color_brewer(palette = "Dark2")
```

## Normalized and Batch Corrected

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
#### CORRECTED

res_pca1 <- mixOmics::pca(t(exprs(corrected)))
res_pca <- cbind(pData(corrected), res_pca1$x)
  
ggplot(res_pca, aes(PC1, PC2, color = Treatment, shape = Batch)) +
  geom_point(size = 3, alpha = 0.8) +
  ggrepel::geom_label_repel(aes(label = rownames(res_pca)), show.legend = F) +
  theme_minimal() +
  xlab(paste0("PC1 (", round(100*(res_pca1$explained_variance)[1], 2), "%)")) +
  ylab(paste0("PC2 (", round(100*(res_pca1$explained_variance)[2], 2), "%)")) +
  scale_color_brewer(palette = "Dark2")
```

# Hierarchical Clustering

## Raw

### By Treatment

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
counts.hc(raw, facs = pData(raw)[, "Treatment", drop = FALSE])
```

### By Batch

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
counts.hc(raw, facs = pData(raw)[, "Batch", drop = FALSE])
```

## Normalized

### By Treatment

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
counts.hc(norm, facs = pData(norm)[, "Treatment", drop = FALSE])
```

### By Batch

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
counts.hc(norm, facs = pData(norm)[, "Batch", drop = FALSE])
```

## Normalized and Batch Corrected

### By Treatment

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
counts.hc(corrected, facs = pData(corrected)[, "Treatment", drop = FALSE])
```

### By Batch

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
counts.hc(corrected, facs = pData(corrected)[, "Batch", drop = FALSE])
```

# Heatmap

## Raw

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
#### RAW

total <- exprs(raw)
target <- pData(raw)

my_group <- as.numeric(as.factor(target$Treatment))
colSide <- brewer.pal(8, "Dark2")[my_group]
colMain <- colorRampPalette( c("green", "black", "red"), space = "rgb")(64)

heatmap(total, ColSideColors = colSide, col = colMain, labRow = NA)
```

## Normalized

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
#### NORMALIZED

total <- exprs(norm)
target <- pData(norm)

my_group <- as.numeric(as.factor(target$Treatment))
colSide <- brewer.pal(8, "Dark2")[my_group]
colMain <- colorRampPalette( c("green", "black", "red"), space = "rgb")(64)

heatmap(total, ColSideColors = colSide, col = colMain, labRow = NA)
```

## Normalized and Batch Corrected

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
#### CORRECTED

total <- exprs(corrected)
target <- pData(corrected)

my_group <- as.numeric(as.factor(target$Treatment))
colSide <- brewer.pal(8, "Dark2")[my_group]
colMain <- colorRampPalette( c("green", "black", "red"), space = "rgb")(64)

heatmap(total, ColSideColors = colSide, col = colMain, labRow = NA)
```

# Important Features

```{r, echo = FALSE, warning = FALSE, comment = NA, message=FALSE}
target <- pData(corrected)
 
###

spc_scatterplot <- function(counts, treat, trans = "log2", minSpC = 2, minLFC = 1){
  
  treat <- as.factor(treat)
  mspc <- data.frame(t(apply(counts, 1, function(x) tapply(x, treat, mean))))
  mx <- apply(mspc[, 1:2], 1, max)
  mn <- apply(mspc[, 1:2], 1, min)
  fl <- apply(mspc[, 1:2], 1, function(x) max(x) >= minSpC) & abs(log2(mx/mn)) >= minLFC
  
  if (trans == "none") {
    
    s_mspc <- mspc[fl ,]
    
    p <- ggplot(mspc, aes(mspc[,1], mspc[,2])) +
      geom_point() +
      geom_point(s_mspc, mapping = aes(x = s_mspc[,1], y = s_mspc[,2], color = "red")) +
      geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
      geom_abline(intercept = 0, slope = 2, linetype = "dashed", color = "navy") +
      geom_abline(intercept = 0, slope = 0.5, linetype = "dashed", color = "navy") +
      xlab(paste("mean SpC(", colnames(mspc)[1], ")", sep = "")) +
      ylab(paste("mean SpC(", colnames(mspc)[2], ")", sep = "")) +
      theme_minimal() +
      ggtitle("Normalized and Batch Corrected") +
      theme(legend.position = "none")
  }
  
  if (trans == "sqrt") {
    
    sqrtm <- sqrt(mspc)
    
    s_mspc <- sqrtm[fl ,]
    
    p <- ggplot(sqrtm, aes(sqrtm[,1], sqrtm[,2])) +
      geom_point() +
      geom_point(s_mspc, mapping = aes(x = s_mspc[,1], y = s_mspc[,2], color = "red")) +
      geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
      geom_abline(intercept = 0, slope = sqrt(2), linetype = "dashed", color = "navy") +
      geom_abline(intercept = 0, slope = 1/sqrt(2), linetype = "dashed", color = "navy") +
      xlab(paste("root mean SpC(", colnames(mspc)[1], ")", sep = "")) +
      ylab(paste("root mean SpC(", colnames(mspc)[2], ")", sep = "")) +
      theme_minimal() +
      ggtitle("Normalized and Batch Corrected") +
      theme(legend.position = "none")
  }
  
  if (trans == "log2") {
    
    logm <- log2(mspc + 0.01)
    mxx <- max(logm[, 1])
    mxy <- max(logm[, 2])
    mx <- max(mxx, mxy)
    
    s_mspc <- logm[fl ,]
    
    p <- ggplot(logm, aes(logm[,1], logm[,2])) +
      geom_point() +
      geom_point(s_mspc, mapping = aes(x = s_mspc[,1], y = s_mspc[,2], color = "red")) +
      geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
      geom_abline(intercept = log2(2), slope = 1, linetype = "dashed", color = "navy") +
      geom_abline(intercept = -log2(2), slope = 1, linetype = "dashed", color = "navy") +
      xlab(paste("log2(mean SpC(", colnames(mspc)[1], "))", sep = "")) +
      ylab(paste("log2(mean SpC(", colnames(mspc)[2], "))", sep = "")) +
      theme_minimal() +
      ggtitle("Normalized and Batch Corrected") +
      theme(legend.position = "none")
  }
  
  return(p)
}

spc_scatterplot(exprs(corrected), target$Treatment, trans = "log2", minSpC = 2, minLFC = 1)
```

